Confirmation of Wide-Field Signatures in Redshifted 21~cm Power Spectra using Murchison Widefield Array Observations

### 2015-05-21 ms.tex
Address ApJL referee comments
From 44209c8:
============

In ms.tex,

(*) Address ApJL referee comments
(*) Add new section "Impact on Instrument Design"

Here we confirm a recently predicted and previously unknown foreground signature in the 3D power spectra of high-redshift 21~cm measurements, wherein the interferometer is sensitive to large-scale structure on all baselines. This is due to the inherently chromatic nature of a wide-field instrument response and is characterized by enhanced power from foreground emission in Fourier modes adjacent to those considered to be most sensitive to the cosmological H~{\sc i} signal. Thus it is a critical input to design and analysis choices of future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array. The simulation which predicted this feature was recently validated against Murchison Widefield Array data but this key element was at or below the noise level. In this paper, we improve the Murchison Widefield Array data sensitivity through coherent averaging of 12 independent snapshots aligned in local sidereal time across different observing nights, and provide the first confirmation of the prediction with a signal-noise ratio $>10$.

@@

A careful design of antenna aperture can significantly mitigate this contamination. Optimal weighting of foreground contaminated Fourier modes may be required to extract the signal with maximum sensitivity. Thus, knowledge of such detailed foreground signatures is key for design and analysis choices of future instruments such as HERA and SKA.

@@

\S\ref{sec:wide-field} is an overview of the role of wide-field measurements in the delay spectral domain and the predicted {\it pitchfork} signature. \S\ref{sec:MWA} describes the analysis of MWA data used to improve the dynamic range of the delay spectra. \S\ref{sec:results} describes the results and confirms the presence of the predicted wide-field effects. \S\ref{sec:summary} summarizes our findings.

@@

The delay power spectrum is defined as \citep[][Paper~I]{par12a}:
\begin{align}\label{eqn:delay-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |\widetilde{V}_b(\tau)|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
with
\begin{align}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi(\frac{\boldsymbol{b}}{\lambda})}{D}, \\
  k_\parallel &\equiv \frac{2\pi\tau\,f_{21}H_0\,E(z)}{c(1+z)^2}, 
\end{align}
where, $A_\textrm{e}$ is the effective area of the antenna, $\Delta B$ is the bandwidth, $\lambda$ is the wavelength of the band center, $k_\textrm{B}$ is the Boltzmann constant, $f_{21}$ is the rest frequency of the 21~cm radiation of H~{\sc i}, $z$ is the redshift, $D\equiv D(z)$ is the transverse comoving distance, $\Delta D$ is the comoving depth along the line of sight, and $h$, $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard cosmology terms. In this paper, we use $\Omega_\textrm{M}=0.27$, $\Omega_\Lambda=0.73$, $\Omega_\textrm{K}=1-\Omega_\textrm{M}-\Omega_\Lambda$, and $H_0=100\,$km$\,$s$^{-1}\,$Mpc$^{-1}$. $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ is in units of K$^2$(Mpc/$h$)$^3$.

@@

The MWA instrument configuration, EoR observing strategy, and analysis procedure applied to individual snapshots used in this study are already described in Paper~I and references therein. In order to reduce thermal fluctuations while maintaining coherence, it is essential to average independent data sets obtained over the same region of sky with identical beamformer settings. Hence, we select a subset of MWA snapshots each of duration 112~seconds obtained over different nights which are aligned to within 72~seconds of each other in {\it local sidereal time} (LST) around a mean LST of 0.04~hours with the MWA tile beam pointed at zenith. The database contains 14 snapshots satisfying these criteria. Two of these snapshots, which contained amplitude and phase artifacts for a significant duration across different baselines, are excluded from our analysis.

The delay spectra of the rest of the snapshots were verified to be coherent in their amplitudes and phases. Results from this coherent averaging are discussed below.

@@

In order to show that low-level ubiquitous diffuse emission is a significant contributor to the {\it pitchfork} signature, contribution from any strong emission from near the Galactic center needs to be minimized. This is best illustrated with northward antenna spacings which map any residual emission from this region (even though the Galactic center has just set over the western horizon) to $\tau\simeq 0$ and thus reduce the impact on higher delay modes (Paper~I). Figure~\ref{fig:3-baseline-comparison-delay-spectra} shows the averaged delay power spectra on three selected baseline vectors oriented northward. Data and noiseless models are shown in black and red respectively. The horizontal dotted black line denotes {\it rms} of thermal fluctuations estimated from data. The vertical dashed line denotes horizon delay limits, and the vertical dotted-dashed lines denote delays at which the responses to coarse band edge flagging are expected.

@@

There is a remarkable agreement in broad morphology between the data and the model. However, some differences in the amplitude scales are noted. We attribute these differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I. At present, these uncertainties prevent a more thorough quantitative comparison between the model and the data. Extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Wayth et al. submitted) to cover the entire hemisphere, and a detailed model of the instrument will improve the accuracy of our model.

@@

@@

Using deeper MWA data, we have confirmed with high significance the earlier prediction that wide-field EoR measurements suffer significant foreground contamination from near the horizon. This has important implications for instrument design and data analysis of future instruments such as HERA and SKA. For instance, SKA plans to use ``stations'' of $\sim$~35~m diameter that act as aperture arrays consisting of pseudo-randomly placed vertical log-periodic dipole arrays. The ``station beam'' will be obtained by a phased addition of the dipole responses. The typical horizon response of these station beams is $\gtrsim$~30~dB lower than at the zenith (private communication from R.~Wayth). HERA will use dishes that limit response near the horizon to $\lesssim -40$~dB, $\sim$~20~dB lower than that of the MWA. Details of both telescopes are being finalized. Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds.

>>>>

Here we confirm a recently predicted and previously unknown foreground signature in the 3D power spectra of high-redshift 21~cm measurements, wherein the interferometer is sensitive to large-scale structure on all baselines. This is due to the inherently chromatic nature of a wide-field instrument response and is characterized by enhanced power from foreground emission in Fourier modes adjacent to those considered to be most sensitive to the cosmological H~{\sc i} signal. Thus it is a critical input to design and analysis choices of future instruments such as the Hydrogen Epoch of Reionization Array and the Square Kilometre Array. The simulation in our recent paper which predicted this feature was validated against Murchison Widefield Array data but this key element was at or below the noise level. In this paper, we improve the Murchison Widefield Array data sensitivity through coherent averaging of 12 independent snapshots aligned in local sidereal time across different observing nights, and provide the first confirmation of the prediction with a signal-noise ratio $>10$.

@@

A careful design of antenna aperture can significantly mitigate this contamination. Optimal weighting of foreground-contaminated Fourier modes may be required to extract the signal with maximum sensitivity. Thus, knowledge of such detailed foreground signatures is key for design and analysis choices of future instruments such as HERA and SKA.

@@

\S\ref{sec:wide-field} is an overview of the role of wide-field measurements in the delay spectral domain and the predicted {\it pitchfork} signature. \S\ref{sec:MWA} describes the analysis of MWA data used to improve the dynamic range of the delay spectra. \S\ref{sec:results} describes the results and confirms the presence of the predicted wide-field effects. \S\ref{sec:impact} underscores their impact on aperture design. \S\ref{sec:summary} summarizes our findings.

@@

The delay power spectrum is defined as \citep[][Paper~I]{par12a}:
\begin{align}\label{eqn:delay-power-spectrum}
  P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel) &\equiv |\widetilde{V}_b(\tau)|^2\left(\frac{A_\textrm{e}}{\lambda^2\Delta B}\right)\left(\frac{D^2\Delta D}{\Delta B}\right)\left(\frac{\lambda^2}{2k_\textrm{B}}\right)^2,
\end{align}
with
\begin{align}
  \boldsymbol{k}_\perp &\equiv \frac{2\pi(\frac{\boldsymbol{b}}{\lambda})}{D}, \\
  k_\parallel &\equiv \frac{2\pi\tau\,f_{21}H_0\,E(z)}{c\,(1+z)^2}, 
\end{align}
where, $A_\textrm{e}$ is the effective area of the antenna, $\Delta B$ is the bandwidth, $\lambda$ is the wavelength of the band center, $k_\textrm{B}$ is the Boltzmann constant, $f_{21}$ is the rest frequency of the 21~cm radiation of H~{\sc i}, $z$ is the redshift, $D\equiv D(z)$ is the transverse comoving distance, $\Delta D$ is the comoving depth along the line of sight, and $h$, $H_0$ and $E(z)\equiv [\Omega_\textrm{M}(1+z)^3+\Omega_\textrm{k}(1+z)^2+\Omega_\Lambda]^{1/2}$ are standard cosmology terms. In this paper, we use $\Omega_\textrm{M}=0.27$, $\Omega_\Lambda=0.73$, $\Omega_\textrm{K}=1-\Omega_\textrm{M}-\Omega_\Lambda$, and $H_0=100\,$km$\,$s$^{-1}\,$Mpc$^{-1}$. $P_\textrm{d}(\boldsymbol{k}_\perp,k_\parallel)$ is in units of K$^2$(Mpc/$h$)$^3$.

@@

The MWA instrument configuration, EoR observing strategy, and analysis procedure applied to individual snapshots used in this study are already described in Paper~I and references therein. {\it Off-zenith} data containing the Galactic center in Paper~I was useful in demonstrating the mapping between delay spectra and sky locations, establishing primary causes of foreground contamination, and devising a technique for mitigating foreground contamination. {\it Pitchfork} signatures which result from low-level ubiquitous diffuse emission are fainter. Since Galactic plane emission, even from directions far away from the primary field of view, can potentially swamp the fainter {\it pitchfork} signatures, zenith pointings which have a maximum avoidance of the Galactic plane are preferred for this study.

To reduce thermal fluctuations while maintaining coherence, it is essential to average independent data sets obtained over the same region of sky with identical instrument settings. Hence, we select a subset of MWA snapshots each of duration 112~seconds obtained over different nights which are aligned to within 72~seconds of each other in {\it local sidereal time} (LST) around a mean LST of 0.04~hours with the MWA tile beam pointed at zenith. The database contains 14 snapshots satisfying these criteria. Two of these snapshots, which contained amplitude and phase artifacts for a significant duration across different baselines, are excluded from our analysis. Results of this coherent averaging are discussed below.

@@

In these observations, the Galactic center is just about to set in the west. Its signature on eastward baselines is seen in the modeled delay spectra (bottom panel) as a marginal brightening of the arm near the negative horizon limit for $|\boldsymbol{b}| < 125$~m, consistent with our findings in Paper~I. This spills over into higher delay modes resulting in the faint ($\lesssim 10^2$~K$^2$~$($Mpc$/h)^3$) vertical stripes at $|\boldsymbol{b}| < 50$~m. The corresponding vertical feature is identified in the averaged data as well.

In order to show that low-level ubiquitous diffuse emission is a significant contributor to the {\it pitchfork} signature, contribution from any strong emission from near the Galactic center needs to be minimized. This is best illustrated with northward antenna spacings which map any residual emission from this region to $\tau\simeq 0$ and thus reduce the impact on higher delay modes (Paper~I). Figure~\ref{fig:3-baseline-comparison-delay-spectra} shows the averaged delay power spectra on three selected baseline vectors oriented northward. Data and noiseless models are shown in black and red respectively. The horizontal dotted black line denotes {\it rms} of thermal fluctuations estimated from data. The vertical dashed line denotes horizon delay limits, and the vertical dotted-dashed lines denote delays at which the responses to coarse band edge flagging are expected.

@@

There is a remarkable agreement in broad morphology between the data and the model. However, some differences in the amplitude scales are noted. For instance, the emission near the positive horizon limit is higher in the data than in the model in both figures. We attribute such differences to uncertainties in the foreground model, the MWA tile power pattern, thermal fluctuations, and other uncertainties noted in Paper~I which limit a more thorough quantitative comparison between the model and the data. In fact, this lays further emphasis on the need for extending the footprint of surveys matching the frequency and angular resolution of observation such as the MWA Commissioning Survey \citep[MWACS;][]{hur14} and the Galactic and Extragalactic MWA Survey (GLEAM; Wayth et al. submitted) to cover the entire hemisphere, and a detailed characterization of the power pattern over the entire hemisphere.

@@

\section{Impact on Instrument Design}\label{sec:impact}

Delay spectrum maps the geometric delays to positions of foreground objects on the sky. Thus the directional power pattern of the antenna has a direct impact on the delay spectrum. Since the contamination in the {\it EoR window} is strongly dependent on sources of emission close to the horizon, design of antenna apertures that suppress their sensitivity towards the horizon is important.

We approached this issue in Paper~I via simulations of different antenna apertures, namely, a dipole, a phased array, and a dish. The relative foreground contamination in the {\it EoR window} was found to be roughly $10^4:10^2:1$ respectively. In addition, the foreground contamination inside the {\it foreground wedge} was found to be suppressed by a factor $10^4$ for a dish relative to an MWA-like phased array thus raising the possibility of accessing the EoR signal in $k$-modes inside the {\it foreground wedge}.

The response of new instruments such as SKA and HERA to foreground contamination has to be investigated thoroughly. With the SKA at low frequencies, plans to deploy ``stations'' of $\sim$~35~m diameter that act as aperture arrays consisting of 256 pseudo-randomly placed vertical log-periodic dipole antennas are being considered. The ``station beam'' will be obtained by a phased addition of the dipole responses. The typical horizon response of these beams is $\gtrsim$~30~dB lower than at zenith\footnote{SKA memo ``Station Response and Imaging Performance of LFAA: 100 MHz to 600 MHz'' by Razavi-Ghods et al.} but the randomness across different stations in the placement of dipoles is expected to reduce the response significantly further as pairs of station voltage beams are multiplied with each other. HERA will use dishes that limit response near the horizon to $\lesssim -40$~dB, $\sim$~20~dB lower than that of the MWA. Their precise beam responses and the resulting foreground contamination, especially near the horizon, will prove to be critical and timely inputs to their aperture designs which are in their final stages.

@@

Using deeper MWA data, we have confirmed with high significance the earlier prediction that wide-field EoR measurements suffer significant foreground contamination from near the horizon. This has important implications for instrument design and data analysis of future instruments such as HERA and SKA. Precise modeling is thus required to gain a complete understanding of the characteristics of the cosmological signal and the foregrounds.
